---
title: "Tree Algorithm using LFSR"
author: "Sue Parkinson"
date: "2022-02-01"
output: workflowr::wflow_html
editor_options:
  chunk_output_type: console
---

## Introduction

First, I load in the code from the `code/fileIO_plotting.R` and `code/lfsr_algorithm.R`.

```{r message=FALSE}
source("code/fileIO_plotting.R")
source("code/lfsr_algorithm.R")
```

## The Data

Load in the simulated test data, and plot a dimensionality reduction.

```{r}
nodetree <- form_tree_from_file('data/NodeTrees/NodeTree4/NodeTree4.csv')
palette=1:20
```

## Algorithm Performance

Compute the drift factorization using LFSR.

```{r}
nodetree$trajectory$lfsr_drift <- lfsr_algorithm(dat=nodetree$matrix,labels=nodetree$labels)
```

## Visualizing the Loadings

When we visualize loadings, I use the function `plot_nodetree_loadings` from `code/fileIO_plotting`. The first plot created by this function is the dimensionality reduction colored according to the loadings for each factor. The second plot shows the actual numerical values of each loading.

Even without back fitting, the loadings do a good job of detecting the tree structure. Unfortunately, L is not binary; that is, the loadings are not constant on each factor.

```{r}
plot_nodetree_loadings(nodetree,
                       loadings=nodetree$trajectory$lfsr_drift$L.pm)
```

## Backfitting

To try to get closer to a true drift factorization, I try to backfit, which by default respects the sparsity pattern in the loadings. Unfortunately, this makes the factorization closer to a naive factorization which is 1 on the loadings for one label, and zero elsewhere. This decreases the ability to interpret these loadings as detecting how populations co-descend from a particular parent node.

```{r}
nodetree$trajectory$lfsr_drift_backfit <- flash.backfit(nodetree$trajectory$lfsr_drift)
plot_nodetree_loadings(nodetree,
                       loadings=nodetree$trajectory$lfsr_drift_backfit$L.pm)
```

## First set loadings to be binary and then backfit?

One sort of silly thing I tried to do is take the sparsity structure defined by the first algorithm, force L to be a binary matrix, and then fit the factors to that. It's sort of a brute force algorithm.

In practice I actually just initialize a new flash object and then use the flash.fix.factors and flash.backfit to initialize with binary loadings

```{r}
#set up priors
driftprior <- as.prior(ebnm.fn=ebnm_point_exponential,sign=1)
Fprior <- ebnm_normal
#initialize with binary L and previous value for F
pm <- nodetree$trajectory$lfsr_drift_backfit$loadings.pm
Linit <- pm[[1]] > 0
Finit <- pm[[2]]
K <- nodetree$trajectory$lfsr_drift_backfit$n.factors
#fit F
nodetree$trajectory$lfsr_binary_loadings <- flash.init(nodetree$matrix) %>%
    flash.set.verbose(0) %>%
    flash.init.factors(list(Linit,Finit),
                       ebnm.fn=c(driftprior,Fprior)) %>%
    flash.fix.factors(kset=1:K,mode=1) %>%
    flash.backfit()
#plot results
plot_nodetree_loadings(nodetree,loadings=nodetree$trajectory$lfsr_binary_loadings$L.pm)
```

It definitely finds the correct structure, but one could picture it may not be entirely robust. This is something to potentially explore.

## Force binary loadings as you go?

Alternatively, you could force the loadings to be binary as you go. This results in roughly the same thing, but perhaps with slighly better results. I changed the functionality of the `add_factor` function to allow for this.

```{r}
nodetree$trajectory$lfsr_drift_force_binary <- lfsr_algorithm(dat=nodetree$matrix,labels=nodetree$labels,allfixed=TRUE)
plot_nodetree_loadings(nodetree,nodetree$trajectory$lfsr_drift_force_binary$L.pm)
```

## Backfit after forcing to be zeros and ones?

Backfitting after enforcing the binary structure doesn't seem to change anything.

```{r}
nodetree$trajectory$lfsr_drift_force_binary_backfit <- flash.backfit(nodetree$trajectory$lfsr_drift_force_binary)
plot_nodetree_loadings(nodetree,nodetree$trajectory$lfsr_drift_force_binary_backfit$L.pm)
```

## Sometimes gives error when all the standard deviations are zero

Unfortunately, I wasn't able to run this algorithm on certain datasets because I kept getting an error about standard deviations being zero. This happened with the nodetree dataset seen above with `lfsr_tol=1e-2` instead of `lfsr_tol=1e-3`, and with the`continuoustree` dataset. However, changing to the internally enforced binary algorithm fixed this.

```{r,eval=FALSE}
#running this code gives an error
res <- lfsr_algorithm(dat=nodetree$matrix,labels=nodetree$labels,lfsr_tol=1e-2,verbose=1)
plot_nodetree_loadings(nodetree,res$L.pm)
```

## Some interesting results on certain datasets

Here are some interesting results using the enforced binary structure.

First is a noisy dataset. We see in these results that with a smaller LFSR tolerance, it detects less structure, as one might anticipate. In some ways the local false sign rate allows us to specify how sure we want to be that we're really finding tree structure. TODO why isn't it working this time?

```{r}
nodetree <- form_tree_from_file('data/NodeTrees/NodeTree3/NodeTree3.csv')
#run method with lfsr 1e-2
# TODO raises standard errors must be positive and nonzero error -- why?
res <- lfsr_algorithm(dat=nodetree$matrix,labels=nodetree$labels,lfsr_tol=1e-1,verbose=1)
plot_nodetree_loadings(nodetree,res$L.pm)
#run method with lfsr 1e-2
res <- lfsr_algorithm(dat=nodetree$matrix,labels=nodetree$labels,lfsr_tol=1e-2)
plot_nodetree_loadings(nodetree,res$L.pm)
#run method with lfsr 1e-3
res <- lfsr_algorithm(dat=nodetree$matrix,labels=nodetree$labels,lfsr_tol=1e-3)
plot_nodetree_loadings(nodetree,res$L.pm)
```

Also, a very simple tree, which it recovers very well.

```{r}
nodetree <- form_tree_from_file('data/NodeTrees/NodeTree1/NodeTree1.csv')

#recreate colormap for this nodetree's labels
color <- nodetree$labels
pal <- colorRamp(palette)
newcolor <- color/max(abs(color)) / 2 + 0.5
newcolor <- alpha(rgb(pal(newcolor)/255),0.4)
#run method
res <- lfsr_algorithm(dat=nodetree$matrix,labels=nodetree$labels,lfsr_tol=1e-3)
plot_nodetree_loadings(nodetree,res$L.pm)
```

## Just using leaf data

Here is the algorithm run on a data set with just leaf data.

```{r}
nodetree <- form_tree_from_file('data/NodeTrees/NodeTree1/NodeTree1.csv')

#recreate colormap for this nodetree's labels
color <- nodetree$labels
pal <- colorRamp(palette)
newcolor <- color/max(abs(color)) / 2 + 0.5
newcolor <- alpha(rgb(pal(newcolor)/255),0.4)
#run method
res <- lfsr_algorithm(dat=nodetree$matrix,labels=nodetree$labels,lfsr_tol=1e-3)
plot_nodetree_loadings(nodetree,res$L.pm)
```

## Does setting L to be binary give the "correct" F?

Lets check by finding the true F and comparing against the computed F.

```{r}
#get computed F
computedF <- res$F.pm
#get true F
labels <- unique(nodetree$labels)
trueF <- matrix(nrow=dim(nodetree$matrix)[2],ncol=length(labels))
## first factor is the mean of the root node
trueF[,1] <- colMeans(nodetree$matrix[nodetree$labels == 0,])
## get the nodes that are parents
parents <- labels[1:(length(labels)/2)]
## initialize index of child and num children per node
childidx <- 2
children_per_node <- 2
## differences are the factors
for (parent in parents){
    parent_mean <- colMeans(nodetree$matrix[nodetree$labels == parent,])
    for (childnum in 1:children_per_node){
        child <- labels[childidx]
        child_mean <- colMeans(nodetree$matrix[nodetree$labels == child,])
        factor <- parent_mean - child_mean
        trueF[,childidx] <- factor
        childidx <- childidx + 1
    }
}
```

First, we should compute the relative error in F. I first rescale both matrices to account for an arbitrary scaling factor.

```{r}
maxabs <- function(data){
    return(max(abs(data)))
}
trueF <- trueF/apply(trueF, 2,max)
computedF <- computedF/apply(computedF, 2,max)
```

```{r}
library('plot.matrix')
par(mar=c(5.1, 4.1, 4.1, 4.1))
plot(trueF,border=NA)
par(mar=c(5.1, 4.1, 4.1, 4.1))
plot(computedF,border=NA)
```

Another sanity check is that the factors should be independent. Below we visualize the true and computed covariance matrices.

```{r}
par(mar=c(5.1, 4.1, 4.1, 4.1))
plot(cov(trueF))
par(mar=c(5.1, 4.1, 4.1, 4.1))
plot(cov(computedF))
```

It looks like they're still fairly correlated, unfortunately. But it wouldn't be hard to get $F$ from $L$ in other ways. I'm just not sure why flash doesn't compute it itself.

## Trying with the covariance matrix instead

This didn't seem to work very well. It throws errors about standard errors being zero in some cases, and in other cases it doesn't detect the tree structure very well.

```{r}
# TODO throws error abt standard errors
# res <- lfsr_algorithm(dat=nodetree$matrix,labels=nodetree$labels,lfsr_tol=1e-2,cov=TRUE)
# plot_nodetree_loadings(nodetree,res$L.pm)

# TODO throws error abt standard errors
# res <- lfsr_algorithm(dat=nodetree$matrix,labels=nodetree$labels,lfsr_tol=1e-1,cov=TRUE)
# plot_nodetree_loadings(nodetree,res$L.pm)

#performs poorly; doesn't separate parents from children very well
res <- lfsr_algorithm(dat=nodetree$matrix,labels=nodetree$labels,lfsr_tol=1e-3,cov=TRUE)
plot_nodetree_loadings(nodetree,res$L.pm)
#performs poorly still
res <- lfsr_algorithm(dat=nodetree$matrix,labels=nodetree$labels,lfsr_tol=1e-4,cov=TRUE)
plot_nodetree_loadings(nodetree,res$L.pm)
```

## Using an exponential prior

TODO

```{r}

```
