---
title: "Tree Algorithm using LFSR"
author: "Sue Parkinson"
date: "2022-02-01"
output: workflowr::wflow_html
editor_options:
  chunk_output_type: console
---

## Introduction

First, I load in the code from the `code/fileIO_plotting.R` and `code/lfsr_algorithm.R`.

```{r message=FALSE}
source("code/fileIO_plotting.R")
source("code/lfsr_algorithm.R")
```

## The Data

Load in the simulated test data, and plot a dimensionality reduction.

```{r}
nodetree <- form_tree_from_file('data/NodeTrees/NodeTree4/NodeTree4.csv')
palette=1:20
```

## Algorithm Performance

Compute the drift factorization using LFSR.

```{r}
nodetree$trajectory$lfsr_drift <- lfsr_algorithm(dat=nodetree$matrix,labels=nodetree$labels)
```

## Visualizing the Loadings

When we visualize loadings, I use the function `plot_nodetree_loadings` from `code/fileIO_plotting`. The first plot created by this function is the dimensionality reduction colored according to the loadings for each factor. The second plot shows the actual numerical values of each loading.

Even without back fitting, the loadings do a good job of detecting the tree structure. Unfortunately, L is not binary; that is, the loadings are not constant on each factor.

```{r}
plot_nodetree_loadings(nodetree,
                       loadings=nodetree$trajectory$lfsr_drift$L.pm)
```

## Backfitting

To try to get closer to a true drift factorization, I try to backfit, which by default respects the sparsity pattern in the loadings. Unfortunately, this makes the factorization closer to a naive factorization which is 1 on the loadings for one label, and zero elsewhere. This decreases the ability to interpret these loadings as detecting how populations co-descend from a particular parent node.

```{r}
nodetree$trajectory$lfsr_drift_backfit <- flash.backfit(nodetree$trajectory$lfsr_drift)
plot_nodetree_loadings(nodetree,
                       loadings=nodetree$trajectory$lfsr_drift_backfit$L.pm)
```

## First set loadings to be binary and then backfit?

One sort of silly thing I tried to do is take the sparsity structure defined by the first algorithm, force L to be a binary matrix, and then fit the factors to that. It's sort of a brute force algorithm.

In practice I actually just initialize a new flash object and then use the flash.fix.factors and flash.backfit to initialize with binary loadings

```{r}
#set up priors
driftprior <- ebnm_point_exponential
Fprior <- ebnm_normal
#initialize with binary L and previous value for F
Linit <- nodetree$trajectory$lfsr_drift_backfit$L.pm > 0
Finit <- nodetree$trajectory$lfsr_drift_backfit$F.pm
K <- nodetree$trajectory$lfsr_drift_backfit$n.factors
#fit F
nodetree$trajectory$lfsr_binary_loadings <- flash.init(nodetree$matrix) %>%
    flash.set.verbose(0) %>%
    flash.init.factors(list(Linit,Finit),
                       ebnm.fn=c(driftprior,Fprior)) %>%
    flash.fix.factors(kset=1:K,mode=1) %>%
    flash.backfit(extrapolate=FALSE)
#plot results
plot_nodetree_loadings(nodetree,loadings=nodetree$trajectory$lfsr_binary_loadings$L.pm)
```

It definitely finds the correct structure, but one could picture it may not be entirely robust. This is something to potentially explore.

## Force binary loadings as you go?

Alternatively, you could force the loadings to be binary as you go. This results in roughly the same thing, but perhaps with slightly better results. I changed the functionality of the `add_factor` function to allow for this. The parameter `allfixed=TRUE` makes it so that all of the values are fixed as you go, not just the zero values in the loadings. In effect, this means that the loadings are forced to be binary as you go.

```{r, eval=FALSE}
# TODO ERROR
nodetree$trajectory$lfsr_drift_force_binary <- lfsr_algorithm(dat=nodetree$matrix,
                                                              labels=nodetree$labels,
                                                              allfixed=TRUE,
                                                              verbose=1)
plot_nodetree_loadings(nodetree,nodetree$trajectory$lfsr_drift_force_binary$L.pm)
```

## Backfit after forcing to be zeros and ones?

Backfitting after enforcing the binary structure doesn't seem to change anything.

```{r, eval=FALSE}
# TODO error bc cell above doesn't work
nodetree$trajectory$lfsr_drift_force_binary_backfit <- flash.backfit(nodetree$trajectory$lfsr_drift_force_binary)
plot_nodetree_loadings(nodetree,nodetree$trajectory$lfsr_drift_force_binary_backfit$L.pm)
```

## Trying Larger LFSR Tolerance Values

By default, I used `lfsr_tol=1e-3`, but this may be unnecessarily stringent. Below, I try using `lfsr_tol = 1e-2`. (This code doesn't work unless `S=1e-12` is used.)

```{r,eval=TRUE}
#not binary
res <- lfsr_algorithm(dat=nodetree$matrix,
                      labels=nodetree$labels,
                      lfsr_tol=1e-2)
plot_nodetree_loadings(nodetree,res$L.pm)
#binary fixed as you go
res <- lfsr_algorithm(dat=nodetree$matrix,
                      labels=nodetree$labels,
                      lfsr_tol=1e-2,
                      allfixed=TRUE)
plot_nodetree_loadings(nodetree,res$L.pm)
```

## Comparison with various `lfsr_tol` on a simple dataset

Here are some interesting results with the binary structure enforced as you go on a fairly noisy dataset. We see in these results that with a smaller LFSR tolerance, it detects less structure. For example, with `lfsr_tol=1e-2`, there are only three factors. With a larger LFSR tolerance of `lfsr_tol=1e-1`, it detects more strucutre, although not 100% perfectly. We see that it finds 9 factors, although factors 4 & 5, 6 & 9, and 7 & 8 are factors ideally would be combined into one.

```{r}
nodetree3 <- form_tree_from_file('data/NodeTrees/NodeTree3/NodeTree3.csv')
#lfsr 1e-1, binary
res <- lfsr_algorithm(dat=nodetree3$matrix,
                      labels=nodetree3$labels,
                      lfsr_tol=1e-1,
                      allfixed=TRUE)
plot_nodetree_loadings(nodetree,res$L.pm)
#lfsr 1e-2, binary
res <- lfsr_algorithm(dat=nodetree3$matrix,
                      labels=nodetree3$labels,
                      lfsr_tol=1e-2,
                      allfixed=TRUE)
plot_nodetree_loadings(nodetree3,res$L.pm)
```

# A very simple tree

It recovers this simple tree with low noise levels fairly well, although it somehow splits the last node up into more levels of complexity than necessary. NOTE: I think this didn't happen with the old versions of `flashier`.

```{r}
nodetree1 <- form_tree_from_file('data/NodeTrees/NodeTree1/NodeTree1.csv')
#run method
res <- lfsr_algorithm(dat=nodetree1$matrix,
                      labels=nodetree1$labels,
                      lfsr_tol=1e-3,
                      allfixed=TRUE)
plot_nodetree_loadings(nodetree1,res$L.pm)
```

## Just using leaf data

Here is the algorithm run on a data set with just leaf data. Unfortunately, it only added one factor. I this is beacuse the LFSR's are numerically close to zero.

```{r}
nodetree1$IsLeaf = nodetree1$csv$IsLeaf=="True"
#run method on leaf data
res_lfsr <- lfsr_algorithm(dat=nodetree1$matrix[nodetree1$IsLeaf,],
                      labels=nodetree1$labels[nodetree1$IsLeaf],
                      lfsr_tol=1e-3)
print(max(res_lfsr$L.lfsr,na.rm = TRUE))
#try divergence fit?
res_div <- div_fit(dat=nodetree1$matrix[nodetree1$IsLeaf,],
                      labels=nodetree1$labels[nodetree1$IsLeaf])
plot_nodetree_loadings(nodetree1,res_div$L.pm) #works great
print(max(res_div$L.lfsr,na.rm = TRUE))
#try drift fit?
res_drift <- drift_fit(dat=nodetree1$matrix[nodetree1$IsLeaf,],
                      labels=nodetree1$labels[nodetree1$IsLeaf])
plot_nodetree_loadings(nodetree1,res_drift$L.pm) #works great
print(max(res_drift$L.lfsr,na.rm = TRUE))
#run method on leaf data with tiny lfsr tol
res_lfsr <- lfsr_algorithm(dat=nodetree1$matrix[nodetree1$IsLeaf,],
                      labels=nodetree1$labels[nodetree1$IsLeaf],
                      lfsr_tol=1e-10)
# TODO I think all the lfsrs are nans bc they're all fixed.
# why is this a non issue otherwise?
print(max(res_lfsr$L.lfsr,na.rm = TRUE)) #all lfsrs are nans...
plot_nodetree_loadings(nodetree1,res_lfsr$L.pm) #works great
```

## Does setting L to be binary give the "correct" F?

Lets check by finding the true F and comparing against the computed F.

```{r}
#get computed F
computedF <- res$F.pm
#get true F
labels <- unique(nodetree1$labels)
trueF <- matrix(nrow=dim(nodetree1$matrix)[2],ncol=length(labels))
## first factor is the mean of the root node
trueF[,1] <- colMeans(nodetree1$matrix[nodetree1$labels == 0,])
## get the nodes that are parents
parents <- labels[1:(length(labels)/2)]
## initialize index of child and num children per node
childidx <- 2
children_per_node <- 2
## differences are the factors
for (parent in parents){
    parent_mean <- colMeans(nodetree1$matrix[nodetree1$labels == parent,])
    for (childnum in 1:children_per_node){
        child <- labels[childidx]
        child_mean <- colMeans(nodetree1$matrix[nodetree1$labels == child,])
        factor <- parent_mean - child_mean
        trueF[,childidx] <- factor
        childidx <- childidx + 1
    }
}
```

We should compute the relative error in F. I first rescale both matrices to account for an arbitrary scaling factor.

```{r}
maxabs <- function(data){
    return(max(abs(data)))
}
trueF <- trueF/apply(trueF, 2,max)
computedF <- computedF/apply(computedF, 2,max)
```

```{r}
library('plot.matrix')
par(mar=c(5.1, 4.1, 4.1, 4.1))
plot(trueF,border=NA)
par(mar=c(5.1, 4.1, 4.1, 4.1))
plot(computedF,border=NA)
```

Another sanity check is that the factors should be independent. Below we visualize the true and computed covariance matrices.

```{r}
par(mar=c(5.1, 4.1, 4.1, 4.1))
plot(cov(trueF))
par(mar=c(5.1, 4.1, 4.1, 4.1))
plot(cov(computedF))
```

It looks like they're still fairly correlated, unfortunately. But it wouldn't be hard to get $F$ from $L$ in other ways. I'm just not sure why flash doesn't compute it itself.

## Trying with the covariance matrix instead

This didn't seem to work very well. It throws errors about standard errors being zero in some cases, and in other cases it doesn't detect the tree structure very well.

```{r}
# TODO throws error abt standard errors
res <- lfsr_algorithm(dat=nodetree$matrix,labels=nodetree$labels,lfsr_tol=1e-2,cov=TRUE)
plot_nodetree_loadings(nodetree,res$L.pm)

# TODO throws error abt standard errors
res <- lfsr_algorithm(dat=nodetree$matrix,labels=nodetree$labels,lfsr_tol=1e-1,cov=TRUE)
plot_nodetree_loadings(nodetree,res$L.pm)

#performs poorly; doesn't separate parents from children very well
res <- lfsr_algorithm(dat=nodetree$matrix,labels=nodetree$labels,lfsr_tol=1e-3,cov=TRUE)
plot_nodetree_loadings(nodetree,res$L.pm)
# TODO throws error abt standard errors
# performs poorly still
res <- lfsr_algorithm(dat=nodetree$matrix,labels=nodetree$labels,lfsr_tol=1e-4,cov=TRUE)
plot_nodetree_loadings(nodetree,res$L.pm)
```
