---
title: "Recursive Partitioning"
author: "Sue Parkinson"
date: "3/28/2022"
output: html_document
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
```

## Inroduction

This file will experiment with using the code by Li et al. as described in ["Hierarchical Community Detection by Recursive Partitioning."](https://doi.org/10.1080/01621459.2020.1833888) and available [here](https://cran.r-project.org/web/packages/HCD/index.html).

## Installing and loading package

```{r,eval=FALSE}
install.packages("HCD")
```
```{r,message=FALSE}
library(HCD)
source("code/fileIO_plotting.R")
```

## Visualizing the covariance matrix

```{r}
nodetree <- form_tree_from_file('data/NodeTrees/NodeTree1/NodeTree1.csv')
image.plot(nodetree$leafcov)
```
## Trying the `HCD` method on various matrices related to the leaf nodes

Doesn't work using the covariance matrix of the nodes

```{r}
res = HCD(nodetree$leafcov)
plot_nodetree(nodetree,color=1+res$labels,palette=1:20,leaf=TRUE)
unique(res$labels)
```

Works using reciprocal of covariance matrix with small values truncated to `1e-5`

```{r}
res = HCD(1/pmax(nodetree$leafcov,1e-5))
plot_nodetree(nodetree,color=1+res$labels,palette=1:20,leaf=TRUE)
paths = res$tree.path
for (spot in 1:length(paths)){
    paths[spot] = substr(paths[spot],1,5)
}
unique(paths)
```

Works with distance matrix.
```{r}
res = HCD(rdist(nodetree$leafmatrix))
plot_nodetree(nodetree,color=1+res$labels,palette=1:20,leaf=TRUE)
paths = res$tree.path
for (spot in 1:length(paths)){
    paths[spot] = substr(paths[spot],1,5)
}
unique(paths)
```

