---
title: "StandardErrorsZero"
author: "Sue Parkinson"
date: "2/11/2022"
output: html_document
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)

library(dplyr)
library(RColorBrewer)
library(scales)
library(flashier)
library(dequer)
library(ebnm)
```

## Reading in the data

First, here is the data. You might have to change the path to data on your local machine.

```{r}
#read in tree data
tree <- vector(mode="list")
tree$csv <- read.csv('data/NodeTrees/NodeTree3/NodeTree3.csv',row.names=1)
tree$matrix <- tree$csv %>%
    select(Raw0:Raw499)
tree$matrix <- as.matrix(tree$matrix)
tree$dimred <- tree$csv %>%
        select(tsne0:tsne1)
```

## Function to visualize the data

In case it is helpful, this is how I've been visualizing this data set. It's a noisy tree with data from internal nodes. The arrows show how the descendancy from the root node to the children. Around each node, samples are normally distributed.

```{r}
#' Plots the dimensionality reduction of a nodetree along with arrows along the branches.
#'
#' @param tree tree object, like the output of form_tree_from_file
#' @param color vector of colors for the data points
#' @param palette color pallete to use for the plot
plot_nodetree <- function(tree,color=1+tree$labels,palette=c("#0000D2","#D2D2D2","#D20000")){
    #create colormap
    ## create color palette function
    pal <- colorRamp(palette)
    ## rescale color to be between 0 and 1, and 0.5 being the "zero" point
    newcolor <- color/max(abs(color)) / 2 + 0.5
    ## use pal and alpha to get rgb value for color
    newcolor <- alpha(rgb(pal(newcolor)/255),0.4)
    ## make colored scatter plot
    plot(tree$dimred,col=newcolor,pch=20)

    # draw arrows between parents and children
    ## get all the different labels of the nodes
    labels <- unique(tree$labels)
    ## get the nodes that are parents
    parents <- labels[1:(length(labels)/2)]
    ## initialize index of child and num children per node
    childidx <- 2
    children_per_node <- 2
    ## draw arrows
    for (parent in parents){
        for (childnum in 1:children_per_node){
            child <- labels[childidx]
            arrows(mean(tree$dimred$tsne0[tree$labels == parent]),
                   mean(tree$dimred$tsne1[tree$labels == parent]),
                   mean(tree$dimred$tsne0[tree$labels == child]),
                   mean(tree$dimred$tsne1[tree$labels == child]),
                   length=0.05)
            childidx <- childidx + 1
        }
    }
}
plot_nodetree(tree,palette=1:7)
```

## Algorithm

Here is the code I've been using that gives the error. The first cell is just some helper functions.

```{r}
#' Helper function that returns the loading from one additional divergence factor.
#' Using the local false sign rate, it sets to zero those loadings for which we cannot be
#' confident of the sign.
#'
#' @param dat the data matrix
#' @param lfsr_tol Tolerance for local false sign rate
#' @param loading binary initial loading to specify known sparsity pattern in the loading (e.g. parental sparsity)
#' @param fl flash object containing the current fit
#' @param divprior divergence prior for loadings
#' @param Fprior prior for the factors. Defaults to a normal prior.
#' @returns the new posterior loading for the additional divergence factor
get_divergence_factor_lfsr <- function(dat,lfsr_tol,loading,fl,divprior,Fprior){
    K <- fl$n.factors
    #initializes factor to the least squares solution
    ls.soln  <- t(crossprod(loading,  dat - fitted(fl))/sum(loading))
    EF <- list(loading, ls.soln)
    next_fl <- fl %>%
        flash.init.factors(
            EF,
            prior.family = c(divprior,Fprior)
        ) %>%
        flash.fix.loadings(kset = K + 1, mode = 1L, is.fixed = (loading == 0)) %>%
        flash.backfit(kset = K + 1)
    loading <- next_fl$loadings.pm[[1]][,K+1]
    lfsr <- next_fl$loadings.lfsr[[1]][,K+1]
    loading[lfsr > lfsr_tol] <- 0
    return(loading)
}

#' Helper function that adds one factor to the flash object.
#'
#' @param dat the data matrix
#' @param loading binary initial loading to specify the sparsity pattern in the loading
#' @param fl flash object to add the factor to
#' @param prior prior for loadings
#' @param Fprior prior for the factors. Defaults to a normal prior.
#' @returns a new flash object with a new factor such that the associated loading matches the
#' sparsity pattern from loading
add_factor <- function(dat,loading,fl,prior,Fprior){
  K <- fl$n.factors
  #initializes factor to the least squares solution
  ls.soln  <- t(crossprod(loading,  dat - fitted(fl))/sum(loading))
  EF <- list(loading, ls.soln)
  #create new flash object
  next_fl <- fl %>%
    flash.init.factors(
      EF,
      prior.family = c(prior,Fprior)
    ) %>%
    flash.fix.loadings(kset = K + 1, mode = 1L, is.fixed = (loading == 0)) %>%
    #only backfit the most recently added factor
    flash.backfit(kset = K + 1)
  return(next_fl)
}

#' Helper function for debugging. Shows how many samples with each label type are activated on
#' a particular loading
#'
#' @param tree tree object; output of form_tree_from_file that has been used in computing a trajectory
#' @param loading the loading in question
get_sums_by_label <- function(tree,loading){
  summary <- as.data.frame(table(tree$labels))
  summary$NumActivated <- 0
  idx <- 1
  for(label in rownames(summary)){
    start_idx <- idx
    idx <- idx + summary[label,"Freq"]
    summary[label,"NumActivated"] <- sum(loading[start_idx:(idx-1)])
  }
  return(summary)
}
```

The next cell is the actual algorithm, which basically uses the local false sign rate to decide when and how to split the data up into hierarchical groups.

```{r}
#' Fits a drift factorization to the data and returns the associated flashier object
#'
#' @param tree tree object; output of form_tree_from_file.
#' @param cov If true, fits to the covariance instead of the data matrix
#' @param divprior prior for intermediate divergence loadings. Defaults to a point Laplace prior.
#' @param driftprior prior for drift loadings. Defaults to a point exponential prior.
#' @param Fprior prior for the factors. Defaults to a normal prior.
#' @param Kmax maximum number of factors to add.
#' @param lfsr_tol Tolerance for local false sign rate
#' @param min_pve If the pve for a factor is less than this tolerance, the factor is rejected
#' @param verbose.lvl The level of verbosity of the function
lfsr_algorithm <- function(tree,
                      cov=FALSE,
                      divprior = prior.point.laplace(),
                      driftprior = as.prior(ebnm.fn = ebnm_point_exponential,sign=1),
                      Fprior = prior.normal(),
                      Kmax = Inf,
                      lfsr_tol = 1e-3,
                      min_pve = 0,
                      verbose.lvl = 0) {
    if (cov){
        dat <- cov(t(tree$matrix))
    }
    else{
        dat <- tree$matrix
    }
    #the first loading will be the all-ones vector
    ones <- matrix(1, nrow = nrow(dat), ncol = 1)
    #first factor will be least sq soln: argmin_f ||Y - ones t(f)||_F^2
    ls.soln <- t(crossprod(ones, dat)/nrow(dat))

    #create flash object with initial drift loading and initial divergence loading
    fl <- flash.init(dat) %>%
        flash.set.verbose(verbose.lvl) %>%
        #initialize L to be the ones vector, and F to be the least squares solution
        flash.init.factors(list(ones, ls.soln),
                           prior.family = c(driftprior,Fprior)) %>%
        #only fixing the first factor, and the we want to fix row loadings, so mode=1
        flash.fix.loadings(kset = 1, mode = 1) %>%
        #backfit to match the priors
        flash.backfit()

    #add first divergence factor to a queue (Breadth-first)
    divergence_queue <- queue()
    new_div <- get_divergence_factor_lfsr(dat,lfsr_tol,loading=ones,fl,divprior,Fprior)
    pushback(divergence_queue,new_div)

    while(length(divergence_queue) > 0 && fl$n.factors < Kmax) {
        if (verbose.lvl > 0) {cat("Length of Queue:",length(divergence_queue),"\n")}
        #pop the first divergence off the queue
        current_divergence <- pop(divergence_queue)
        #split into positive and negative parts
        splus <- matrix(1L * (current_divergence > 0), ncol = 1)
        if (sum(splus) > 0 && fl$n.factors < Kmax) {
            print("HERE")
            if (verbose.lvl > 0) {print(get_sums_by_label(tree,splus))}
            print("HERE 2")
            #add drift loading
            next_fl <- add_factor(dat,splus,fl,driftprior,Fprior)
            if (next_fl$pve[next_fl$n.factors] > min_pve){
                fl <- next_fl
                #enqueue new divergence
                new_div <- get_divergence_factor_lfsr(dat,lfsr_tol,loading=splus,fl,divprior,Fprior)
                pushback(divergence_queue,new_div)
            }
        }
        sminus <- matrix(1L * (current_divergence < 0), ncol = 1)
        if (sum(sminus) > 0 && fl$n.factors < Kmax) {
            if (verbose.lvl > 0) {print(get_sums_by_label(tree,sminus))}
            #add drift loading
            next_fl <- add_factor(dat,sminus,fl,driftprior,Fprior)
            if (next_fl$pve[next_fl$n.factors] > min_pve){
                fl <- next_fl
                #enqueue new divergence
                new_div <- get_divergence_factor_lfsr(dat,lfsr_tol,loading=sminus,fl,divprior,Fprior)
                pushback(divergence_queue,new_div)
            }
        }
        if (verbose.lvl > 0) {cat("Factors:", fl$n.factors, "\n")}
    }

    return(fl)
}
```

Finally, running the data on this algorithm gives an error. This is set to `eval=FALSE` right now, but the error it gives is

```{r,eval=FALSE}
res <- lfsr_algorithm(tree,lfsr_tol=1e-2,cov=TRUE,verbose=1)

```
